Subsidiary Recursion in Coq

### 2022-01-14 paper/paper.tex
working on Section 4 (implementation of Subrec)

Definition inList : ListF List -> List := inn ListF.
Definition mkNil : List := inList Nil.
Definition mkCons (hd : A) (tl : List) : List := inList (Cons hd tl).
Definition toList : list A -> List.
Definition fromList : List -> list A.
\end{verbatim}
  \caption{Some basics from \texttt{List.v}, specializing the functor-generic derivation of subsidiary recursion to lists (\texttt{List.v})}
  \label{fig:listf}
\end{figure}

@@

  This file specializes the development in \verb|Subrec.v| to the case
  of lists (parametrized by the type \verb|A| of elements).  In
  general, to use our development to get subsidiary recursion over
  some datatype, one will have a similar ``shim'' file.  The file
  defines the signature functor \verb|ListF|, shown in
  Figure~\ref{fig:listf}.  Using \verb|Subrec|, we then get a type
  \verb|List|.  This is not to be confused with the type \verb|list|
  of lists in Coq's standard library.  As noted previously, our
  development is meant to be used in extension of existing inductive
  datatypes, not replacing them.  The figure also shows constructors
  \verb|mkNil| and \verb|mkCons| for \verb|List|, and types for
  conversion functions between \verb|List| and \verb|list| (see
  Section~\ref{sec:deriv} for the code).

@@

We can write this combinatory using our infrastructure for subsidiary recursion.
For this to work, we need to supply the mapped function with the fold function
for \verb|mapThrough|'s recursion.  This is so that the mapped function can
initiate a subsidiary recursion, returning a value in the abstract type \verb|R|
of \verb|mapThrough|'s recursion.  So the type we will use for mapped functions
is:
\begin{verbatim}
Definition mappedT(A B : Set) : Set :=
  forall(R : Set)(fo:FoldT (Alg (ListF A)) R), A -> R -> B * R.
\end{verbatim}
\noindent This type is more informative than the Haskell type,
since it shows that the second component of the returned value
must have type \verb|R|, and hence must be (hereditarily) a tail
of the input of type \verb|R|.

@@

Using \verb|mapThrough|, we can write a quite concise function for
\emph{run-length encoding}, a basic data-compression algorithm where
maximal sequences of $n$ occurrences of element $e$ are summarized by
the pair $(n,e)$~\cite{datacomp}.  In Haskell, invoking \verb|span| and
\verb|mapThrough| (defined above), the code is simply
\begin{verbatim}
rle :: Eq a => [a] -> [(Int,a)]
rle = mapThrough compressSpan
  where compressSpan a as =
          let (p,s) = span (== a) as in
            ((1 + length p, a),s)
\end{verbatim} 
\noindent (Recall that \verb|(== a)| is a Haskell \emph{section}
testing its input for equality with \verb|a|.)  The
\verb|compressSpan| helper function gathers up all elements at the
start of the tail \verb|as| that are equal to the head \verb|a|.  This
prefix is returned as \verb|p|, with the remaining suffix as \verb|s|.
The pair \verb|(1 + length p, a)| is returned to summarize
\verb|a :: p|.  We then use \verb|mapThrough| to iterate
\verb|compressSpan| through the suffix \verb|s|.

@@

@@

As we have seen, our definitions require a form of positive-recursive
types, to allow algebras to accept fold functions that themselves
require algebras, and also for the definition of \verb|Subrec|.  But
as recalled already, full positive-recursive types are incompatible
with Coq's type theory~\cite{coquand88}.  It is worth noting that one
can impose some restrictions on large eliminations which then allow
positive-recursive types~\cite{blanqui05}.  This approach would
require changing the underlying theory.  To avoid this, we here take a
different approach, exploiting Coq's impredicative polymorphism.

@@

Noncanonicity leads to some issues, as we turn next to the problem of
inductive reasoning about subsidiary recursions.  With some care,
however, we can avoid pitfalls, leaving us with a form of
positive-recursive type that enables our definitions to go through.

@@

Mendler introduced the basic idea of using universal abstraction to
support compositional termination checking; an accessible source
is~\cite{mendler91}.  This recursor has type
\[
\all{X}{(\all{R}{(R \to X) \to F\ R \to X}) \to \mu\ F\ \to X}
\]
\noindent We have adopted this idea to the constructor of the type
\verb|Mu| (Section~\ref{sec:mu}).  Previous work explored the
categorical perspective on Mendler-style recursion~\cite{uustalu99}.
Others have explored the possibility of using it with negative type
schemes~\cite{ahn11}.

>>>>

Definition List := Subrec ListF .
Definition inList : ListF List -> List := inn ListF.
Definition mkNil : List := inList Nil.
Definition mkCons (hd : A) (tl : List) : List := inList (Cons hd tl).
Definition toList : list A -> List.
Definition fromList : List -> list A.
\end{verbatim}
  \caption{Some basics from \texttt{List.v}, specializing the functor-generic derivation of subsidiary recursion to lists (\texttt{List.v})}
  \label{fig:listf}
\end{figure}

@@

  This file specializes the development in \verb|Subrec.v| to the case
  of lists (parametrized by the type \verb|A| of elements).  In
  general, to use our development to get subsidiary recursion over
  some datatype, one will have a similar ``shim'' file.  The file
  defines the signature functor \verb|ListF|, shown in
  Figure~\ref{fig:listf}.  We then define \verb|List| to be
  \verb|Subrec|, with the instantiation of \verb|F| to \verb|ListF A|.
  This type \verb|List| is not to be confused with the type
  \verb|list| of lists in Coq's standard library.  As noted
  previously, our development is meant to be used in extension of
  existing inductive datatypes, not replacing them.  The figure also
  shows constructors \verb|mkNil| and \verb|mkCons| for \verb|List|,
  and types for conversion functions between \verb|List| and
  \verb|list| (see Section~\ref{sec:deriv} for the code).

@@

We can write this combinator using our infrastructure for subsidiary recursion.
For this to work, we need to supply the mapped function with the fold function
for \verb|mapThrough|'s recursion.  This is so that the mapped function can
initiate a subsidiary recursion, returning a value in the abstract type \verb|R|
of \verb|mapThrough|'s recursion.  So the type we will use for mapped functions
is:
\begin{verbatim}
Definition mappedT(A B : Set) : Set :=
  forall(R : Set)(fo:FoldT (Alg (ListF A)) R), A -> R -> B * R.
\end{verbatim}
\noindent This type is more informative than the Haskell type,
since it shows that the second component of the returned value
must have type \verb|R|, and hence must be (hereditarily) a tail
of the input of type \verb|R|.

@@

Using \verb|mapThrough|, we can quite concisely implement
\emph{run-length encoding}, a basic data-compression algorithm where
maximal sequences of $n$ occurrences of element $e$ are summarized by
the pair $(n,e)$~\cite{datacomp}.  In Haskell, invoking \verb|span| and
\verb|mapThrough| (defined above), the code is simply
\begin{verbatim}
rle :: Eq a => [a] -> [(Int,a)]
rle = mapThrough compressSpan
  where compressSpan a as =
          let (p,s) = span (== a) as in
            ((1 + length p, a),s)
\end{verbatim} 
\noindent (Recall that \verb|(== a)| is a Haskell \emph{section}
testing its input for equality with \verb|a|.)  The
\verb|compressSpan| helper function gathers up all elements at the
start of the tail \verb|as| that are equal to the head \verb|a|.  This
prefix is returned as \verb|p|, with the remaining suffix as \verb|s|.
The pair \verb|(1 + length p, a)| is returned to summarize
\verb|a :: p|.  We then use \verb|mapThrough| to iterate
\verb|compressSpan| through the suffix \verb|s|.

@@

Let us now consider the implementation of the
interface we have used for the preceding examples.
The first step is our weakened form of positive-recursive
types.

@@

As we have seen, our definitions require a form of positive-recursive
types, to allow algebras to accept fold functions that themselves
require algebras, and also for the definition of \verb|Subrec| (which
we will see in more detail in the next section).  As already noted,
full positive-recursive types are incompatible with Coq's type
theory~\cite{coquand88}.  One can impose some restrictions on large
eliminations which then enable positive-recursive
types~\cite{blanqui05}.  This approach would require changing the
underlying theory.  To avoid this, we here take a different approach,
exploiting Coq's impredicative polymorphism.

@@

Noncanonicity must be handled carefully when reasoning about functions
defined with our interface.  We will see examples in
Section~\ref{sec:examplesi}.  First, though, let us complete the
exposition of our implementation of subsidiary recursion.

\subsection{The implementation of \texttt{Subrec} (\texttt{Subrec.v})}

The type \verb|Subrec| is defined in Figure~\ref{fig:subrec}, as a
fixed-point of \verb|SubrecF : Set -> Set|.  We take this fixed-point
with \verb|Mu|, discussed in the previous section, and obtain \verb|roll|
and \verb|unroll| functions between \verb|SubrecF Subrec| and \verb|Subrec|.
Unrolling \verb|Subrec| gives us the type
\begin{verbatim}
forall (X : Set -> Set) (FunX : Functor X), Alg Subrec X -> X Subrec
\end{verbatim}
\noindent So we see that \verb|Subrec| is the type of functions which,
for all algebras with anchor type \verb|Subrec| and functorial carrier
\verb|X|, compute a value of type \verb|X Subrec|.  This is a generalization
of the functor-generic type for the Church encoding:
\[
\forall\ X.\ \textit{Alg}\ X \to X
\]
\noindent where $\textit{Alg}\ X$ is $F\ X \to X$.  
We elide the implementation of the \verb|roll| and \verb|unroll| functions, but
note that \verb|unroll| makes use of functoriality of carriers \verb|X|.

\begin{figure}
\begin{verbatim}
Definition SubrecF(C : Set) := 
  forall (X : Set -> Set) (FunX : Functor X), Alg C X -> X C.
Definition Subrec := Mu SubrecF.
Definition roll: SubrecF Subrec -> Subrec.
Definition unroll: Subrec -> SubrecF Subrec.
\end{verbatim}
\caption{Definition of \texttt{Subrec} as a fixed-point of \texttt{SubrecF}}
\label{fig:subrec}
\end{figure}

The rest of the interface for \verb|Subrec| is shown in
Figure~\ref{fig:subrecb}.  We have \verb|fold|, which is a fold
function with anchor type \verb|Subrec|.  To fold an algebra
\verb|alg| with carrier \verb|X| (with \verb|fmap| function given by
\verb|FunX|) over \verb|d : Subrec|, \verb|fold| \verb|unroll|s the
definition of \verb|Subrec| and applies that to the algebra (with its
carrier).

More interesting is the definition of \verb|inn|.

\begin{figure}
\begin{verbatim}
Definition fold : FoldT Alg Subrec := 
  fun X FunX alg d => unroll d X FunX alg.

Definition inn : F Subrec -> Subrec :=
  fun d => roll (fun X xmap alg =>
                 unrollAlg alg Subrec (fun x => x) fold (fold X xmap alg) d).

Definition out{R:Set}(fo:FoldT Alg R) : R -> F R :=
  fo F FunF (rollAlg (fun _ _ _ _ d => d)).
\end{verbatim}
\caption{The rest of the interface for \texttt{Subrec}}
\label{fig:subrecb}
\end{figure}

@@

Mendler introduced the basic idea of using universal abstraction to
support compositional termination checking; an accessible source
is~\cite{mendler91}.  He introduces a functor-generic recursor of type
\[
\all{X}{(\all{R}{(R \to X) \to F\ R \to X}) \to \mu\ F\ \to X}
\]
\noindent We have adopted this idea to the constructor of the type
\verb|Mu| (Section~\ref{sec:mu}).  Previous work explored the
categorical perspective on Mendler-style recursion~\cite{uustalu99}.
Others have explored the possibility of using it with negative type
schemes~\cite{ahn11}.
